Rectangulotopes

### 2024-04-04 draft/rectangulotopes.tex
shuffled intro

% Loday and anti-Loday accents (?)
\makeatletter
\newcommand{\oset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits^{
    \vbox to#1{\kern-2\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
%
\newcommand{\uset}[3][0ex]{%
  \mathrel{\mathop{#3}\limits_{
    \vbox to#1{\kern-7\ex@
    \hbox{$\scriptstyle#2$}\vss}}}}
\makeatother
%
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
\newcommand{\upArc}[1]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}}}
\newcommand{\downArc}[1]{\smash{\raisebox{-.05cm}{$\oset[.2ex]{#1}{\smile}$}}}
% yin and yang accents (?)
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yinArc}[2]{\smash{\raisebox{.05cm}{$\uset[0ex]{#1}{\frown}$}\hspace{-.3ex}\raisebox{-.03cm}{$\oset[.2ex]{#2}{\smile}$}}}
\newcommand{\yangArc}[2]{\smash{\raisebox{-.03cm}{$\oset[.2ex]{#1}{\smile}$}\hspace{-.3ex}\raisebox{.05cm}{$\uset[0ex]{#2}{\frown}$}}}
% weak and strong equivalence on S_n
\newcommand{\weakeq}{\asymp}
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}% % J: I would like to stack \backsim and \sim instead

@@

The goal of this paper is to describe the quotientopes of congruences of the weak Bruhat order whose classes are in bijection with (equivalence classes of) \defn{rectangulations}, defined as decompositions of a square into finitely many axis-aligned rectangles.
The combinatorics of rectangulations has been studied for decades~\cite{MR2233287,MR2871762,MR2864445,MR2763051,MR3084577,MR3878132,MR4598046}, and the order-theoretic take on these structures has already proved useful.
%, in particular in the context of electronic engineering applications~\cite{}.
\vincent{More citations on electronic engineering...}
There are two relevant lattice congruences related with rectangulations.
%
On the one hand, the \defn{weak rectangulation congruence} (or \defn{Baxter congruence}), studied by S.~Law and N.~Reading~\cite{MR2871762}, and whose classes are one-to-one with \defn{weak rectangulations} (or \defn{diagonal rectangulations}), with \defn{twin binary trees}~\cite{Giraudo}, and with \defn{Baxter permutations}. \vincent{Give more Baxter references here.}
The corresponding quotientopes will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$.
The skeleton of $\WRP(n)$ is isomorphic to the flip graph of weak rectangulations of size $n$ defined by Law and Reading~\cite{MR2871762}.
%
On the other hand, the \defn{strong rectangulation congruence}, whose classes are in bijection with \defn{strong rectangulations}~\cite{MR2864445}.
The latter is mentioned by N.~Reading~\cite{MR3335492} (see for instance in Theorem 1.2, and Example 4.11), and studied more recently by A.~Asinowski, J.~Cardinal, S.~Felsner, and É.~Fusy~\cite{ACFF24}.
The corresponding quotientopes will be referred to as the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.
The skeleton of $\SRP(n)$ is isomorphic to the flip graph of strong rectangulations of size $n$ defined by Meehan~\cite{MR3697823} (see also~\cite{ACFF24}).

Rectangulotopes enrich the family of natural quotientopes, alongside associahedra, permutreehedra~\cite{MR3856522}, certain brick polytopes~\cite{PilaudSantos-brickPolytopes, PilaudStump-brickPolytopes, Pilaud-brickAlgebra}, and certain graphical zonotopes~\cite{Pilaud-brickAlgebra, Pilaud-acyclicReorientationLattices}, providing instructive examples of application of the theory of lattice congruences to concrete combinatorial objects.
Our main results are elementary vertex and facet descriptions of both families of quotientopes.
For vertices, we describe the coordinates of the vertex corresponding to a weak or strong rectangulation with simple product formulas in the spirit of J.-L.~Loday's realization in~\cref{thm:loday}.
As mentioned in~\cite{PilaudSantosZiegler}, there is no such formula for arbitrary quotientopes, which makes rectangulotopes join the restricted club of Loday type quotientopes, jointly with associahedra and permutreehedra~\cite{MR3856522}.
For facets, we show that both weak and strong rectangulotopes have $2^n-2$ facets (the maximal possible number of facets of a deformed permutahedron), and provide simple formulas for the right hand side of the inequality corresponding to a given proper nonempty subset of~$[n]$.
We now give the necessary definitions for precisely stating these vertex and facet descriptions of the weak and strong rectangulotopes.

@@

\begin{itemize}
\item Terminology: rectangles, \defn{vertices}, \defn{segments}, \defn{generic} rectangulation
\item \defn{Wall slides}
\item Two equivalence classes: \defn{weak equivalence} preserving rectangle-segment adjacencies, and \defn{strong equivalence} preserving rectangle-rectangle adjacencies.
\item Weak equivalence classes have a \defn{diagonal} representative.
\item \defn{Flip graphs} on rectangulations
\end{itemize}

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
We denote by $i \, \backslash \, j$ the situation where the rectangle labeled $i$ touches the rectangle labeled $j$ either from the left or from below in $R$.
We also use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise.

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote nonempty and consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

The \defn{Minkowski sum} of two polytopes~$\polytope{P}$ and~$\polytope{Q}$ is~$\polytope{P} + \polytope{Q} \eqdef \set{\b{p} + \b{q}}{\b{p} \in \polytope{P} \text{ and } \b{q} \in \polytope{Q}}$.
Recall that for any vector~$\b{v} \ne \b{0}$, the face of~$\polytope{P} + \polytope{Q}$ maximizing the scalar product with~$\b{v}$ is the Minkowski sum of the faces of~$\polytope{P}$ and~$\polytope{Q}$ maximizing the scalar product with~$\b{v}$.
In particular, if~$\b{v}$ is a generic direction, the vertex of~$\polytope{P} + \polytope{Q}$ extremal in direction~$\b{v}$ is just the sum of the vertices of~$\polytope{P}$ and~$\polytope{Q}$ that are extremal in direction~$\b{v}$.

@@

@@

Given a rectangulation~$R$, the source and target trees~$S(R)$ and~$T(R)$ define two partial orders on $[n]$ (the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top).
The \defn{weak poset}~$P_w(R)=([n],\prec_w)$ of the rectangulation~$R$ is the transitive closure of the union of the two partial orders defined by the intertwining binary trees~$S(R)$ and~$T(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, whose minimum is the root of~$S(R)$ and whose maximum is the root of~$T(R)$.
We define the set~$\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

\begin{theorem}[\cite{LawReading}]
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of each congruence class are the twisted Baxter permutations,
  \item the maximal elements are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

\subsection{Up and down arcs}

@@

These arcs can be classified into two families of arcs: the up arcs and the down arcs.
For an interval $I$ of $[n]$ of size at least~$2$, the \defn{up arc}~$\upArc{I}$ (resp.~\defn{down arc}~$\downArc{I}$) is the arc that starts at~$\min(I)$ and ends at~$\max(I)$, and passes above (resp.~below) all remaining elements of~$I$, that is
\begin{align*}
\upArc{I} & \eqdef \big( \min(I), \max(I), I \ssm \{\min(I), \max(I)\}, \varnothing \big) \\
\qquad\text{and}\qquad
\downArc{I} & \eqdef \big( \min(I), \max(I), \varnothing, I \ssm \{\min(I), \max(I)\} \big).
\end{align*}
Note that the basic arc~$(i, i+1, \varnothing, \varnothing)$ is both an up and down arc.

@@

Therefore, the quotientope is realized by a Minkowski sum of the shard polytopes of the up arcs and the down arcs.
These quotientopes are very basic ones.

\begin{lemma}
  \label{lem:lodaysp}
%  Given an up arc~$\alpha$ defined by an interval $I$, the up shard polytope $\SP(\alpha)$ is a translate of the standard simplex~$\conv \set{ \b{e}_i }{ i\in I }$.
  The up shard polytope $\SP(\upArc{I})$ is a translate of the simplex~$\conv \set{ \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
%  Given a down arc~$\alpha$ defined by an interval $I$, the down shard polytope $\SP(\alpha)$ is a translate of the negative standard simplex~$\conv \set{ - \b{e}_i }{ i\in I }$.
  The down shard polytope $\SP(\downArc{I})$ is a translate of the negative simplex~$\conv \set{ - \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:weakMinkowski}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all up and down shard polytopes.
  In particular, it is the Minkowski sum of two opposite associahedra of J.-L.~Loday~\cite{LawReading}.
\end{lemma}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f_{\weakeq}(S) \eqdef \#\bigset{ I\text{ interval of } [n] }{ I \not\subseteq S \text{ and } I \not\subseteq [n]\ssm S }.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

%The proof relies on the following simple lemma stating that the submodular function of a Minkowski sum of deformed permutahedra is the sum of the submodular function of the summands.
%
%\begin{lemma}
%  \label{lem:submodsum}
%  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
%\end{lemma}

\begin{proof}
  %From \cref{lem:weakMinkowski,lem:submodsum},
  The submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the up and down shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. The shard polytope~$\SP(\upArc{I})$ of an up arc~$\upArc{I}$ is given by the submodular function
  \[
  f_{\upArc{I}}(S) \eqdef \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, the shard polytope~$\SP(\downArc{I})$ of a down arc~$\downArc{I}$ is given by the submodular function
  \[
  f_{\downArc{I}}(S) \eqdef - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f_{\weakeq}(S) = \sum_{I\text{ interval of }[n]} f_{\upArc{I}}(S) + f_{\downArc{I}}(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals contained neither in~$S$ nor in its complement.
\end{proof}

@@

For every weak rectangulation $R$, we wish to give the coordinates of a point $\b{p}(R)\in\R^n$ such that $\WRP(n)$ is the convex hull of the points~$\b{p}(R)$ for all weak rectangulations~$R$.
Recall that in a Minkowski sum of polytopes, the vertex that is extremal in a generic direction is the sum of the vertices of the summands that are extremal in this direction.
We therefore need to understand which vertices of the translates of the shard polytopes defined in \cref{lem:lodaysp,lem:antilodaysp} are extremal for a direction given by a permutation.

\begin{lemma}
  \label{lem:lodaymax}
  Let $\pi \in \f{S}_n$ and $I$ be a subset of~$[n]$.
  The unique vertex of the up shard polytope~$\SP(\upArc{I})$ (resp.~down shard polytope~$\SP(\downArc{I})$) that is extreme with respect to the direction $\pi^{-1}$ is $\b{e}_i$, where $i \eqdef \arg\max_{j\in I} \pi^{-1}(j)$ (resp.~where~$i \eqdef \arg\min_{j\in I} \pi^{-1}(j)$).
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on~$n$ rectangles, with source and target trees~$S$ and~$T$, and let $\b{p}(R)$ the vertex of $\WRP(n)$ associated with $R$.
  Given $i\in [n]$, the number of up arcs~$\upArc{I}$ such that $\b{e}_i$ is the extremal vertex of $\SP(\upArc{I})$ contributing to~$\b{p}(R)$ is
  \[
  \loday{w}_i =  h^{T}_i\cdot v^{T}_i,
  \]
   where $h^T_i$ and $v^T_i$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
\end{lemma}

@@

\begin{theorem}[\cite{Reading}]
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item the congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of each congruence class are the 2-clumped permutations,
  \item the maximal elements are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}

>>>>

% Loday and anti-Loday accents (?)
\newcommand{\loday}[1]{\smash{\overset{\frown}{#1}}}
\newcommand{\antiloday}[1]{\smash{\overset{\smile}{#1}}}
% yin and yang accents (?)
\newcommand{\yang}[1]{\smash{\overset{\backsim}{#1}}}
\newcommand{\yin}[1]{\smash{\overset{\sim}{#1}}}
% weak and strong equivalence on S_n
\newcommand{
\weakeq}{\asymp}
%\makeatletter
%\DeclareRobustCommand{\strongeq}{%
%  \mathbin{\mathpalette\scaletoplus@{\mkern1mu \smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}} \mkern1mu}}%
%}
%\newcommand{\scaletoplus@}[2]{%
%  \vcenter{%
%    \hbox{\sbox\z@{$\m@th#1+$}\resizebox{\wd\z@}{!}{$\m@th#1#2$}}%
%  }%
%}
%\makeatother
\newcommand{\strongeq}{\mathbin{\smash{\begin{smallmatrix} \backsim \\[-.3cm] \sim \end{smallmatrix}}}}% % J: I would like to stack \backsim and \sim instead

@@

The goal of this paper is to describe the quotientopes of congruences of the weak Bruhat order whose classes are in bijection with equivalence classes of rectangulations, defined as decompositions of a square into rectangles.

@@

The combinatorics of rectangulations has been studied for decades~\cite{MR2233287,MR2871762,MR2864445,MR2763051,MR3084577,MR3878132,MR4598046}.
Interestingly, many combinatorial results were initially motivated by applications to the design of \defn{floorplans} for very large scale integrated circuits, and published in electrical engineering journals~\cite{}.
The recent order-theoretic take on these structures has proved fruitful.

We define a \defn{rectangulation} of size $n$ as a decomposition of the square into $n$ axis-parallel rectangles with disjoint interiors.
The \defn{segments} of a rectangulation are the inclusionwise maximal line segments composed of edges of the rectangles, excluding the edges of the decomposed square.
We suppose throughout that no four rectangles have a common vertex, hence that the rectangulations are \defn{generic} in that sense.
In order to focus on the combinatorial structure of a rectangulation, we introduce two equivalence relations between rectangulations.
First, we say that a rectangle $r$ is \defn{above} another rectangle $s$ (and $s$ is \defn{below} $r$) if either the bottom edge of $r$ lies in the same segment as the top edge of $s$, or if $r$ is above another rectangle $t$ that is above $s$.
Similarly, a rectangle $r$ is \defn{on the left of} another rectangle $s$ (and $s$ is \defn{on the right of} $r$) if either the right edge of $r$ lies in the same segment as the left edge of $s$, or if $r$ is on the left of another rectangle $t$ that is on the left of $s$.

Two rectangulations are said to be \defn{weakly equivalent} if there exists a bijection between their rectangles that preserves the above-below and left-right relations between the rectangles.
Weak equivalence classes of rectangulations will be referred to as \defn{weak rectangulations}.
(Note that this is a slight, yet convenient abuse of terminology, as weak rectangulations are really sets of rectangulations.)
On the other hand, two rectangulations are said to be \defn{strongly equivalent} when there exists a bijection between their rectangles that not only preserves the above-below and left-right relations, but also the adjacency relation between rectangles.
We will naturally refer to strong equivalence classes of rectangulations as \defn{strong rectangulations}.

In order to make sense of these definitions, it is useful to consider \defn{wall slides} in a rectangulation: the local changes consisting of shifting a horizontal segment vertically, or a vertical segment horizontally, while extending or shortening the incident segments accordingly.
Performing a wall slide in a rectangulation $R$ leads to a rectangulation that is always weakly equivalent to $R$.
However, if the wall slide changes the adjacency relation between the rectangles, then the resulting rectangulation is not strongly equivalent to $R$ anymore.
This is illustrated in Figure~\ref{fig:rectequiv}.
A \defn{diagonal rectangulation} is a rectangulation in which every rectangle intersects the top-left to bottom-right diagonal of the square.
It is simple to check that every weak rectangulation has a diagonal representative, hence weak rectangulations can also be thought of as diagonal rectangulations.

\begin{figure}
  \caption{\label{fig:rectequiv}Weak and strong equivalence between rectangulations.}
\end{figure}

\subsection{Weak and strong rectangulotopes}

Weak and strong rectangulations of size $n$ have been shown to define congruences of the weak Bruhat order on $\f{S}_n$.
The \defn{weak rectangulation congruence} (or \defn{Baxter congruence}) was first explicitly studied by S.~Law and N.~Reading~\cite{MR2871762}.
Its classes are one-to-one with weak rectangulations, but also with \defn{twin binary trees}~\cite{Giraudo}, and with \defn{Baxter permutations}~\cite{}. % more baxter refs
The corresponding quotientopes will be referred to as the \defn{weak rectangulotopes} and denoted by $\WRP(n)$.
On the other hand, the \defn{strong rectangulation congruence}, whose classes are in bijection with strong rectangulations~\cite{MR2864445}.
The latter is mentioned by N.~Reading~\cite{MR3335492} (see for instance in Theorem 1.2, and Example 4.11), and studied more recently by A.~Asinowski, J.~Cardinal, S.~Felsner, and É.~Fusy~\cite{ACFF24}.
The corresponding quotientopes will be referred to as the \defn{strong rectangulotopes} and denoted by $\SRP(n)$.

An important motivation for studying these quotientopes is the informative structure of their skeleta, which are \defn{flip graphs} on the rectangulations.
A flip is a local, reversible change in the structure of the rectangulation.
Flips in rectangulations come in three flavors:
\begin{description}
\item[Simple flips] replace a vertical segment incident to exactly two rectangles by a horizontal segment, and vice-versa.
\item[Pivoting] operations transform a pair of adjacent rectangles whose union is not a rectangle, changing a left-right pair into an above-below pair, and vice-versa.
\item[Wall slide] flips are wall slides that modify the adjacency of the rectangles, removing exactly one adjacency and adding exactly one. These flips are only relevant in strong rectangulations.
\end{description}

The various types of flips are illustrated in Figure~\ref{fig:flips}.
We refer to S.~Law and N.~Reading~\cite{MR2871762}, J.~Cardinal, V.~Sacrist\'an, and R.~Silveira~\cite{}, E.~Meehan~\cite{MR3697823}, and A.~Asinowski, J.~Cardinal, S.~Felsner, and \'E.~Fusy~\cite{ACFF24} for detailed descriptions of the flip graphs on both weak and strong rectangulations.
Just like associahedra encode the rotation graphs on binary trees, the rectangulotopes directly yield flip graphs on rectangulations.

\begin{lemma}
  The skeleton of $\WRP(n)$ is isomorphic to the flip graph on weak rectangulations of size $n$.
  Similarly, the skeleton of $\SRP(n)$ is isomorphic to the flip graph on strong rectangulations of size $n$.
\end{lemma}

Rectangulotopes enrich the family of natural quotientopes, alongside associahedra, permutreehedra~\cite{MR3856522}, certain brick polytopes~\cite{PilaudSantos-brickPolytopes, PilaudStump-brickPolytopes, Pilaud-brickAlgebra}, and certain graphical zonotopes~\cite{Pilaud-brickAlgebra, Pilaud-acyclicReorientationLattices}, providing instructive examples of application of the theory of lattice congruences to concrete combinatorial objects.
Our main results are elementary vertex and facet descriptions of both families of quotientopes.
For vertices, we describe the coordinates of the vertex corresponding to a weak or strong rectangulation with simple product formulas in the spirit of J.-L.~Loday's realization in~\cref{thm:loday}.
As mentioned in~\cite{PilaudSantosZiegler}, there is no such formula for arbitrary quotientopes, which makes rectangulotopes join the restricted club of Loday type quotientopes, jointly with associahedra and permutreehedra~\cite{MR3856522}.
For facets, we show that both weak and strong rectangulotopes have $2^n-2$ facets (the maximal possible number of facets of a deformed permutahedron), and provide simple formulas for the right hand side of the inequality corresponding to a given proper nonempty subset of~$[n]$.

We now give the necessary definitions for precisely stating these vertex and facet descriptions of the weak and strong rectangulotopes.

@@

In order to exhibit a similar realization for the strong rectangulotopes, we need to count subsets of leaves that lie in some subtrees of both the source and target trees of a rectangulation.
Two leaves of the trees $T(R)$ and $S(R)$ are said to be \defn{common leaves} if the edges to their parents lie on the same segment of $R$.
We denote by %$i|j$ and $\frac{\ j\ }i$
$i \, \backslash \, j$ the situation where the rectangle labeled $i$ touches the rectangle labeled $j$ either from the left or from below in $R$. %, respectively.
We also use the Iverson bracket $\llbracket \varphi\rrbracket$, which is equal to 1 if $\varphi$ holds, and 0 otherwise.

\begin{theorem}
  The $(n-1)$-dimensional strong rectangulotope is realized by the polytope~$\SRP (n)$ defined equivalently as
  \begin{enumerate}[(i)]
  \item convex hull of the points
  \[
  \sum_{i,j\in [n], i< j} (\yin{w}^R_{i,j} - \yang{w}^R_{i,j})\cdot (\b{e}_i - \b{e}_j),
  \]
   for all strong rectangulations $R$ on $n$ rectangles, with
  \[
%  \begin{split}
%    \yin{w}_{i,j} \eqdef & h(T, i) \cdot cv (T, S, i, j)\cdot h(S, j)\cdot \llbracket \neg i\backslash j \rrbracket \\% \text { if not } i\backslash j, \text{and } 0\text{ otherwise.} \\ % \neg i|j \text{ and } \neg \frac{\ j\ }i ] \\
%    \yang{w}_{i,j} \eqdef & v(S, i) \cdot ch (S, T, i, j)\cdot v(T, j) \cdot \llbracket \neg j\backslash i \rrbracket,
%  \end{split}
    \yin{w}^R_{i,j} \eqdef h^T_i \cdot cv^{T,S}_{i,j}\cdot h^S_ j\cdot \llbracket \neg \, i \, \backslash \, j \rrbracket
    \qquad\text{and}\qquad
    \yang{w}^R_{i,j} \eqdef v^S_i \cdot ch^{S,T}_{i,j}\cdot v^T_j \cdot \llbracket \neg \, j \, \backslash \, i \rrbracket,
  \]
 where:
  \begin{itemize}
  \item $S$ and $T$ are the source and target trees of the rectangulation~$R$,
  \item $h^T_i$ and $v^T_i$ denote the number of leaves in the horizontal and vertical subtrees of $i$ in~$T$,
  \item $ch^{T,S}_{i,j}$ (resp.~$cv^{T,S}_{i,j}$) denote the number of common leaves of the horizontal (resp.~vertical) subtree of $i$ in $T$ and the horizontal (resp.~vertical) subtree of $j$ in $S$.
  \end{itemize}
  \item the intersection of the hyperplane defined by~$\sum_{i \in [n]} x_i = \binom{n+1}{2}$ with the halfspaces defined by~$\sum_{i \in S} x_i \le \#\set{I,J}{I \not\subseteq S \text{ and } J \not\subseteq [n] \ssm S} + \#\set{I,J}{I \not\subseteq [n] \ssm S \text{ and } J \not\subseteq S}$ for all subsets~$\varnothing \ne S \subsetneq [n]$, where~$I,J$ in the sets denote consecutive intervals of~$[n]$.
  \end{enumerate}
\end{theorem}

@@

@@

\begin{lemma}
  \label{lem:submodsum}
  Consider a collection $\polytope{P}_1, \polytope{P}_2,\ldots ,\polytope{P}_k$ of deformed permutahedra defined by the submodular functions $f_1,f_2,\ldots ,f_k$. Then the Minkowski sum $\sum_i \polytope{P}_i$ is defined by the submodular function $f \eqdef \sum_i f_i$.
\end{lemma}

@@

Given a rectangulation $R$, the two trees $T(R)$ and $S(R)$, where the horizontal edges are oriented from left to right and the vertical edges are oriented from bottom to top,
define two partial orders on $[n]$.
We define the \defn{weak poset} $P_w(R)=([n],\prec_w)$ of a rectangulation $R$ as the transitive closure of the union of the two partial orders defined by the intertwining binary trees $T(R)$ and $S(R)$~\cite{MR4014603}.
The weak poset is a two-dimensional lattice, the minimum of which is the root of $S(R)$ and the maximum is the root of $T(R)$.
We define the set $\mathcal{L}(P_w(R))$ as the linear extensions of the weak poset:
\[
\mathcal{L}(P_w(R)) \eqdef \bigset{\pi\in\f{S}_n }{ i\prec_w j\implies \pi^{-1}(i) < \pi^{-1}(j)}.
\]

\begin{theorem}
  The subsets $\mathcal{L}(P_w(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $
\weakeq$ on the weak Bruhat order on $\f{S}_n$.
  Furthermore:
  \begin{itemize}
  \item The congruence classes are one-to-one with weak rectangulations,  
  \item the minimal elements of each congruence class are the twisted Baxter permutations,
  \item the maximal elements are the co-twisted Baxter permutations,
  \item the congruence classes each contain a single Baxter permutation, and are one-to-one with Baxter permutations.
  \end{itemize}
\end{theorem}

@@

\subsection{up and down arcs}

@@

% jean: From this, do we directly have a description of the face lattice of $\WRP(n)$?}

@@

Therefore, the quotientope is realized by a Minkowski sum of the shard polytopes defined by two families of arcs: the \defn{up arcs} and the \defn{down arcs}, that go respectively above and below all elements of an interval $I$ of $[n]$.

\begin{lemma}
  \label{lem:lodaysp}
  Given a up arc $A$ defined by an interval $I$, the up shard polytope $\SP(A)$ is a translate of the standard simplex $\conv \set{ \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:antilodaysp}
  Given a down arc $A$ defined by an interval $I$, the down shard polytope $\SP(A)$ is a translate of the negative standard simplex $\conv \set{ - \b{e}_i }{ i\in I }$.
\end{lemma}

\begin{lemma}
  \label{lem:weakMinkowski}
  The weak rectangulotope $\WRP(n)$ is realized by the Minkowski sum of all up and down shard polytopes.
  In particular, it is the Minkowski sum of two opposite Loday associahedra.
\end{lemma}

\vincent{Say also the minkowski sum of all faces of the standard simplex corresponding to intervals, and all their opposite.}
\vincent{Say that this was already known by Law and Reading.}

@@

\begin{lemma}
  The weak rectangulotope $\WRP(n)$ is realized by the following submodular function:
  \[
  f(S) \eqdef \#\bigset{ I\text{ interval of } [n] }{ I\cap S\not=\emptyset \text{ and } I\cap ([n]\ssm S) \not= \emptyset }.
  \]
  %that maps $S$ to the number of intervals $I$ of $[n]$ such that $I\cap S\not=\emptyset$ and $I\not\subseteq S$. 
\end{lemma}

\begin{proof}
  From \cref{lem:weakMinkowski,lem:submodsum}, the submodular function defining the realization of $\WRP(n)$ is the sum of the submodular functions defining the up and down shard polytope in \cref{lem:lodaysp,lem:antilodaysp}. For a up arc $A$ defined by an interval $I$ of $[n]$, we have that its shard polytope $\SP(A)$ is realized by the submodular function
  \[
  \loday{f}_I(S) \eqdef \llbracket S\cap I\not=\emptyset \rrbracket .
  \]
  (Recall that $\llbracket \varphi\rrbracket$ is equal to 1 if $\varphi$ holds, and 0 otherwise.)
  Similarly, for a down arc $A$ defined by an interval $I$, the shard polytope $\SP(A)$ is defined by the following submodular function:
  \[
  \antiloday{f}_I(S) \eqdef - \llbracket I\subseteq S \rrbracket .
  \]
  Their sum
  \[
  f(S) \eqdef \sum_{I\text{ interval of }[n]} \loday{f}_I(S) + \antiloday{f}_I(S)
  \]
  is therefore exactly the number of intervals of $I$ that intersect $S$ but are not contained in $S$, hence the number of intervals intersecting both $S$ and its complement.
\end{proof}

@@

For every rectangulation $R$, we wish to give the coordinates of a point $p(R)\in\R^n$ such that $\WRP(n)$ is the convex hull of the points $p(R)$ for all rectangulations $R$.
Recall that in a Minkowski sum of polytopes, the vertex that is extremal in a generic direction is the sum of the vertices of the summands that are extremal in this direction.
We therefore need to understand which vertices of the translates of the shard polytopes defined in \cref{lem:lodaysp,lem:antilodaysp} are extremal for a direction given by a permutation.

\begin{lemma}
  \label{lem:lodaymax}
  Let $\pi\in\f{S}_n$ and consider a up arc $A$ defined by an interval $I$.
  The unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $\b{e}_i$, where $i \eqdef \arg\max_{j\in I} \pi^{-1}(j)$.
  Similarly, if $A$ is a down arc defined by $I$, the unique vertex of $\SP(A)$ that is extreme with respect to the direction $\pi^{-1}$
  is $-\b{e}_i$, where $i \eqdef \arg\min_{j\in I} \pi^{-1}(j)$.
\end{lemma}

@@

\begin{lemma}
  Let $R$ be a rectangulation on $n$ rectangles and $p(R)$ the vertex of $\WRP(n)$ associated with $R$.
  Given $i\in [n]$, the number of up arcs $A$ such that $\b{e}_i$ is the extremal vertex of $\SP(A)$ contributing to $p(R)$ is
  \[
  \loday{w}_i =  h(T(R), i)\cdot v(T(R),i),
  \]
   where $h(T,i)$ and $v(T,i)$ denote respectively the number of leaves in the horizontal and vertical subtrees of $i$ in the tree $T$.
\end{lemma}

@@

\begin{theorem}
  The subsets $\mathcal{L}(P_s(R))$ of $\f{S}_n$, for all rectangulations $R$ on $n$ rectangles, 
  are equivalence classes of a congruence relation $\strongeq$ on the weak Bruhat order on $\f{S}_n$.
    Furthermore:
  \begin{itemize}
  \item The congruence classes are one-to-one with strong rectangulations,  
  \item the minimal elements of each congruence class are the 2-clumped permutations,
  \item the maximal elements are the co-2-clumped permutations.
  \end{itemize}
\end{theorem}
